---
title: "Single-Cell RNA-seq Analysis of Adipocyte Differentiation"
subtitle: "Data from Ravikanth Nanduri et al. (2022)"
author: "Abu Saadat"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
  cache: false
---

# Introduction

This script replicates the analysis performed in **Ravikanth Nanduri et al. (2022)**. this study examined the role of HMGNs in white adipocyte browning by comparing wild-type (WT) mice and cells to genetically derived mice and cells lacking the two major members of the HMGN protein family, HMGN1 and HMGN2 (DKO
mice).
The experimental design included WT vs DKO mice, 2 replicates each, at two time points, 0 and 6 days. At day 0, cells were in a preadipocyte state, while at day 6 they had differentiated into adipocytes. Each time point had two replicates of each condition (WT vs DKO), for a total of 8 samples.

## Setup

```{r}

setwd("/run/media/saadat/A/Trainig/GettingStarted_scRNASeq/")

library(Seurat)
library(tidyverse)
library(miQC)
library(scater)
library(SingleCellExperiment)
library(presto)
library(glmGamPoi)
library(celldex)
library(SingleR)
library(monocle3)
library(SeuratWrappers)

```

## Load Data

```{r}

#List files 
files <- list.files(path="data/",recursive=T,pattern="*.h5")

#Create a list of count matrices
h5_read <- lapply(paste0("data/",
                         files), Read10X_h5)

#Assign names to each list element based on file names
names(h5_read)<- sapply(files, 
                        function(x){str_split_1(x,"_")[1]},
                        USE.NAMES = FALSE)

#Create Seurat objects for each sample

# All samples
adp <- mapply(CreateSeuratObject,counts=h5_read,  
              project=names(h5_read),
              MoreArgs = list(min.cells = 3, min.features = 200))

#Merge all samples into a single Seurat object
adp <- merge(adp[[1]], y = adp[2:length(adp)], 
                  add.cell.ids = names(adp),project="Adipose")

rm(h5_read)

```

## Adding Metadata

```{r}

adp$condition <- ifelse(str_detect(adp@meta.data$orig.ident, "^W"),
                        "WT","DKO")
adp$time_point <- ifelse(str_detect(adp@meta.data$orig.ident, "0"),
                         "Day 0","Day 6")
adp$cond_tp <- paste(adp$condition, adp$time_point)

```

```{r}

# Visualization the number of cell counts per sample
adp@meta.data %>% 
  ggplot(aes(x=orig.ident, fill=orig.ident)) + 
  geom_bar(color="black") +
  stat_count(geom = "text", colour = "black", size = 3.5, 
            aes(label = ..count..),
            position=position_stack(vjust=0.5))+
  theme_classic() +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("Number of Cells per Sample")

```

# Quality Control

## MiQC Model Fitting

miQC is an adaptive scRNA-seq quality-control method that flags low-quality cells by jointly modeling two common QC signals, the fraction of reads mapping to mitochondrial genes and the number of detected genes—using a probabilistic mixture-model framework, then filtering cells based on each cell’s posterior probability of being “compromised” rather than using fixed, hand-tuned cutoffs

```{r}

mt_genes <- grepl("^mt-",  rownames(adp))
feature_ctrls <- list(mito = rownames(adp)[mt_genes])

# Converting Seurat object to SingleCellExperiment object (handle v5 assay with multiple layers)

DefaultAssay(adp) <- "RNA"

obj <- JoinLayers(adp)
sce <- as.SingleCellExperiment(obj)

# Running QC metrics
sce <- scuttle::addPerCellQC(sce, subsets = feature_ctrls)
plotMetrics(sce)

# MiQC model fitting
model <- mixtureModel(sce)
summary(model)

plotModel(sce, model)

# Plotting Filtering Results
plotFiltering(sce, model)
sce <- filterCells(sce, model)

# transfer QC metrics back to Seurat
qc_df <- as.data.frame(SummarizedExperiment::colData(sce))
adp <- Seurat::AddMetaData(adp, metadata = qc_df)
adp_filt <- subset(adp, cells = rownames(qc_df)[sce$prob_compromised < 0.75])

```

## Normalization and Scaling

SCTransform is a per-gene regression model that turns raw UMI counts into something close to “variance-stabilized, depth-corrected” expression values.

Each gene is modeled using a negative binomial generalized linear model (GLM) where the sequencing depth (i.e., total UMI count per cell) is used as a covariate. The Pearson residuals from this model are then scaled to have unit variance, resulting in normalized expression values that can be used for downstream analysis.

```{r}

adp_norm <- SCTransform(adp_filt, 
                        method = "glmGamPoi",  # Improved speed of learning
                        vst.flavor = "v2",     # Updated regularization
                        vars.to.regress = "percent.mt",
                        conserve.memory = TRUE, # To reduce memory usage
                        verbose = TRUE)

# adp_norm <- readRDS("/run/media/saadat/A/Trainig/GettingStarted_scRNASeq/outputs/adp_norm_mt.rds")

```

## Dimensionality Reduction 

```{r}

adp_norm <- RunPCA(adp_norm, verbose = FALSE, assay = "SCT")

# dimensions 1 to 9
DimHeatmap(adp_norm, dims = 1:9, cells = 500, balanced = TRUE,ncol=3)

# Elbow Plot
ElbowPlot(adp_norm,ndims = 50)

```

## Clustering

```{r}

# Nearest Neighbor Graph
adp_norm <- FindNeighbors(adp_norm, dims = 1:20)
# Clustering
adp_norm <- FindClusters(adp_norm, algorithm = 4, resolution = 0.1, random.seed = 111)
# UMAP Visualization
adp_norm <- RunUMAP(adp_norm, dims = 1:20)

DimPlot(adp_norm, reduction = "umap", group.by = c("orig.ident", "seurat_clusters","time_point","condition","cond_tp"),
        alpha=0.4, ncol=2)

```

## Finding cluster markers

```{r}

# Prepping for marker finding
adp_norm<-PrepSCTFindMarkers(adp_norm,verbose=T)

Idents(adp_norm) = "SCT_snn_res.0.1"
DefaultAssay(adp_norm) = "SCT"

# Finding markers for all clusters
adp_norm_markers <- FindAllMarkers(adp_norm, only.pos = TRUE)

#ordering the results
adp_norm_markers <- adp_norm_markers %>% 
  arrange(cluster,desc(avg_log2FC), desc(p_val_adj))

#examine a small subset
adp_norm_markers %>%
    group_by(cluster) %>%
    slice_max(n = 5, order_by = avg_log2FC)

# adp_norm <- readRDS("/run/media/saadat/A/Trainig/GettingStarted_scRNASeq/outputs/adp_norm_Cls.rds")
```

## Visualizing Marker Expression

```{r}

top20<-adp_norm_markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 20) %>%
    ungroup() 
DoHeatmap(adp_norm, features = top20$gene, slot="scale.data") + NoLegend()

```

## Comparing Time Points


```{r}

Idents(adp_norm) = "time_point"
table(Idents(adp_norm))

# Finding differentially expressed genes between Day 0 and Day 6

TP_de <- FindMarkers(adp_norm,ident.1="Day 6",ident.2="Day 0",test.use="wilcox",
                     min.pct = 0.2, only.pos = TRUE)

DimPlot(adp_norm,group.by="time_point", alpha=0.4)
FeaturePlot(adp_norm,features= c("Acta2", "A2m"),order=T)
```

# Pseudo-bulk Differential Expression Analysis

```{r}

pseudo_adp=AggregateExpression(adp_norm,assays="RNA", return.seurat=T, group.by=c("orig.ident","time_point","condition","cond_tp"))

head(pseudo_adp@assays$RNA$counts)

# Rename cell names
pseudo_adp = RenameCells(pseudo_adp,new.names=gsub("_.*","",pseudo_adp$orig.ident))
pseudo_adp$orig.ident=gsub("_.*","",pseudo_adp$orig.ident)

# Differential Expression between Day 0 and Day 6 in pseudo-bulked data
Idents(pseudo_adp)="time_point"
bulk_adp_de = FindMarkers(pseudo_adp, ident.1="Day 6", ident.2="Day 0", test.use="DESeq2")

head(bulk_adp_de)

# Visualizing Top Markers per Cluster
Idents(adp_norm)="SCT_snn_res.0.1"
DotPlot(adp_norm,features= rownames(bulk_adp_de)[1:10],dot.scale = 3)+coord_flip()

```

# Cell Annotation

###  Annotation by Differential Expressed Gene Markers

From the paper where this data was obtained, the following list of gene markers was obtained:
Mmp3    : preadipocytes
Mki67   : proliferating cells
Fabp4   : differentiating beige adipocytes and differentiated beige adipocytes
Scd1    : differentiated beige adipocytes
Ucp1    : differentiated beige adipocytes
Ppargc1a: differentiated beige adipocytes
Elovl3  : differentiated beige adipocytes
Cidea   : differentiated beige adipocytes

```{r}

markers=c("Mmp3","Mki67","Fabp4","Scd1","Ucp1","Ppargc1a","Elovl3","Cidea")
FeaturePlot(adp_norm,features= markers,ncol=3,order=T)

```

## Annotation using SingleR

```{r}

adp.sce = as.SingleCellExperiment(adp_norm,assay="SCT") #This selects *only* the SCT assay
mouseRNASeq = celldex::MouseRNAseqData()
table(mouseRNASeq$label.main)

# common <- intersect(rownames(adp.sce), rownames(mouseRNASeq))
# common <- common[seq_len(min(2000, length(common)))] 

# Running SingleR
annot = SingleR(test=adp.sce, ref=mouseRNASeq, labels=mouseRNASeq$label.main)
# Adding annotations to Seurat object
table(annot$pruned.labels, useNA="ifany") 

# Adding annotations to Seurat object
adp_norm$mouseRNASeq.main = annot$pruned.labels

# Visualizing Annotations
DimPlot(adp_norm,group.by="mouseRNASeq.main")

```

## Trajectory Analysis


```{r}

# Visualize the subset
DimPlot(adp_norm, cells.highlight = colnames(adipocytes), 
        cols.highlight = "red", cols = "gray90") +
  ggtitle("Highlighted: Adipocyte Subset") 

# Extracting Adipocytes
adipocytes <- subset(adp_norm, subset = mouseRNASeq.main == "Adipocytes")

# Reprocessing Adipocyte Subset

DefaultAssay(adipocytes) <- "SCT"

adipocytes <- RunPCA(adipocytes, verbose = FALSE)

# Check PCs
ElbowPlot(adipocytes, ndims = 30)
DimHeatmap(adipocytes, dims = 1:9, cells = 500, balanced = TRUE, ncol = 3)

# Re-run UMAP with fewer dimensions (since it's a subset)
adipocytes <- RunUMAP(adipocytes, dims = 1:15, verbose = FALSE)

# Re-cluster at higher resolution to capture sub-populations
adipocytes <- FindNeighbors(adipocytes, dims = 1:10)
adipocytes <- FindClusters(adipocytes, resolution = 0.3, random.seed = 111)

# Visualize the subset
DimPlot(adipocytes, group.by = c("seurat_clusters", "time_point", "condition"),
        ncol = 3, pt.size = 1)

```

### Monocle3

```{r}

# Convert subset to cell_data_set
cds_adp <- as.cell_data_set(adipocytes)

# Transfer UMAP from Seurat
cds_adp@int_colData@listData$reducedDims$UMAP <- 
  adipocytes@reductions$umap@cell.embeddings

# Set up clusters
cds_adp@clusters$UMAP$clusters <- adipocytes$seurat_clusters

# Create named partitions vector
partitions <- rep(1, ncol(cds_adp))
names(partitions) <- colnames(cds_adp)
cds_adp@clusters$UMAP$partitions <- partitions


# Learn trajectory graph
cds_adp <- learn_graph(cds_adp, use_partition = FALSE)

# Visualize trajectory
plot_cells(cds_adp,
           color_cells_by = "cluster",
           label_groups_by_cluster = TRUE,
           label_leaves = TRUE,
           label_branch_points = TRUE,
           graph_label_size = 3,
           cell_size = 1.5)

# Color by time point
plot_cells(cds_adp,
           color_cells_by = "time_point",
           label_cell_groups = FALSE,
           label_leaves = FALSE,
           label_branch_points = TRUE,
           cell_size = 1.5)

# Color by condition (WT vs DKO)
plot_cells(cds_adp,
           color_cells_by = "condition",
           label_cell_groups = FALSE,
           label_leaves = FALSE,
           label_branch_points = TRUE,
           cell_size = 1.5)

```

## Ordering Cells in Pseudotime

```{r}

# Select root cells from Day 0 (preadipocytes)
root_cells <- colnames(cds_adp)[which(colData(cds_adp)$time_point == "Day 0")]

# Order cells
cds_adp <- order_cells(cds_adp, root_cells = root_cells)

# Plot pseudotime
plot_cells(cds_adp,
           color_cells_by = "pseudotime",
           label_cell_groups = FALSE,
           label_leaves = FALSE,
           label_branch_points = TRUE,
           cell_size = 1.5)

```

### Identifying Genes that Change with Pseudotime

```{r}
# Fit models to identify genes that change with pseudotime
trajectory_genes <- graph_test(cds_adp, neighbor_graph = "principal_graph")

# Ensure rowData has gene_short_name (CRITICAL FIX)
if(!"gene_short_name" %in% colnames(rowData(cds_adp))) {
  rowData(cds_adp)$gene_short_name <- rownames(cds_adp)
}
# Get top genes
top_trajectory_genes <- trajectory_genes %>%
  filter(q_value < 0.05) %>%
  arrange(desc(morans_I)) %>%
  head(50) %>%
  rownames_to_column(var = "gene_short_name")

# Plot top trajectory genes
plot_cells(cds_adp,
           genes = head(top_trajectory_genes$gene_short_name, 5),
           show_trajectory_graph = TRUE,
           label_cell_groups = FALSE,
           label_leaves = FALSE)

```
### Marker Expression over Pseudotime

```{r}
adipocyte_markers <- c("Mmp3", "Mki67", "Fabp4", "Ucp1", "Cidea", "Ppargc1a")

plot_cells(cds_adp,
           genes = adipocyte_markers,
           show_trajectory_graph = TRUE,
           label_cell_groups = FALSE)
```
